---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bis620.2022

<!-- badges: start -->
[![R-CMD-check](https://github.com/nokkvidan/bis620.2022/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/nokkvidan/bis620.2022/actions/workflows/R-CMD-check.yaml)
[![Test coverage](https://github.com/nokkvidan/bis620.2022/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/nokkvidan/bis620.2022/actions/workflows/test-coverage.yaml)
[![Lint](https://github.com/nokkvidan/bis620.2022/actions/workflows/lint.yaml/badge.svg)](https://github.com/nokkvidan/bis620.2022/actions/workflows/lint.yaml)
[![Codecov](https://codecov.io/gh/nokkvidan/bis620.2022/branch/main/graph/badge.svg?token=U4USARWKRM)](https://codecov.io/gh/nokkvidan/bis620.2022)
<!-- badges: end -->

# Motivations

It is estimated that 1 in 10 people in America have diabetes and 1 in 5 of those with the health condition have not been clinically diagnosed [1]. Type 1 diabetes occurs when the body is unable to produce insulin and is inherited from their family, but type 2 diabetes occurs when the body is unable to use insulin properly and can be prevented or delayed with improvements to a patient's lifestyle. 

Our goal is to identify what kinds of health indicators might be linked with Type 2 diabetes and pre-diabetes.

By using this project, healthcare professionals will be able to quickly determine the likelihood of a patient's susceptibility to diabetes, allowing for earlier detection and providing guidelines to inform lifestyle changes and treatments. This can ultimately improve individual patient outcomes and reduce the overall proportion of diabetes in the population.

Additionally, this project can help researchers study the significant correlation between specific health indicators and the likelihood of developing diabetes. By using the functions provided in this project, researchers can quickly analyze the relationship between different variables and diabetes, potentially uncovering new insights into the causes and risk factors of the condition.

To accomplish this goal, we developed this diabetes risk prediction R package which helps facilitate the analysis of health indicators such as age, BMI, and sex. Through a number of machine learning algorithms we can analyze a large Kaggle data set of patients with known diabetes status and health indicators [2]. Theses algorithms use a combination of continuous and categorical variables to create a model that can predict the likelihood of a patient having diabetes. The following models are available in this package as functions: generalized linear model, XGBoost and random forests. These functions also include additional options for optimization and evaluation.

The data set `diabetes` has been attached to the package and is available to use as demonstrated in the example section below.

### Badge overview

[Coverage page](https://app.codecov.io/gh/nokkvidan/bis620.2022)

[Lint results](https://github.com/nokkvidan/bis620.2022/actions/workflows/lint.yaml)

## Installation

You can install the development version of bis620.2022 from [GitHub](https://github.com/nokkvidan/bis620.2022) with:

``` r
# install.packages("devtools")
devtools::install_github("nokkvidan/bis620.2022")
```

## Usage

To use this package, you must first load it using the following command:

```{r library}
library(bis620.2022)
```

Once the package is loaded, you can use the function `data(diabetes)` to get the diabetes data provided with the package. 

The package includes the following functions:

* `vis_2vars()`: Visualize any two variables in the dataset.
* `vis_dist()`: Visualize the distributions of the non categorical variables in the data set
* `vis_num()`: Visualize the non categorical variables in the dataset with boxplots. Shows both scaled and non scaled variables.
* `glm_model()`: Fits a GLM to the input data and returns the model object.
* `boost_model()`: Fits an XGBoost model to the input data and returns the model object.
* `rf_model()`: Fits a random forest model to the input data and returns the model object.

# Examples
```{r}
data(diabetes)
vis_2vars(df, 'Diabetes_binary', 'BMI')
```


# Contributing

We welcome contributions to the `bis620.2022` package. If you would like to contribute, please follow these steps:

1. Fork the repository and clone it to your local machine.
2. Create a new branch for your changes.
3. Make your changes and push them to your fork.
4. Open a pull request and describe your changes.

# License

The `bis620.2022` package is (not) licensed under the MIT License.

# Citations 

[1]  “The Facts, Stats, and Impacts of Diabetes.” CDC, Centers of Disease Control and Prevention, https://www.cdc.gov/diabetes/library/spotlights/diabetes-facts-stats.html. 

[2] “Diabetes Health Indicators Data Set.” Kaggle, https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset. 
